Resolvable Ambiguity

### 2019-07-05 paper/paper.tex
Add rigor to section 4
Adds definition of interpreted language. Streamlines the definition
of an ambiguous word, as well as the definitions of resolvably
ambiguous words and languages.

\newcommand{\support}{\mathit{Supp}}
\newcommand{\NT}{V} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{L} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}
\newcommand{\regex}{\mathit{Reg}}
\newcommand{\reqpl}{(}
\newcommand{\reqpr}{)}
\newcommand{\reqp}[1]{\reqpl#1\reqpr}
\newcommand{\pospl}{[}
\newcommand{\pospr}{]}
\newcommand{\posp}[1]{\pospl#1\pospr}
\newcommand{\dfa}{\mathit{dfa}} % Go from label to its corresponding dfa
\newcommand{\passthrough}{\mathit{passthrough}} % Go from label to a set of (label x bool), where each element denotes a child production that can be an only child (i.e., no other terminals whatsoever), and the bool states whether it requires parentheses
\newcommand{\nt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs

@@

Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense. We will call this meaning an \emph{interpretation} of a word. Note that a single word may have multiple interpretations. As such, we define a language as:

\begin{itemize}
\item An alphabet $\T$.
\item A set of interpretations $T$.
\item A function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
\end{itemize}

\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees. In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs in Fig.~\ref{fig:arith-example}:

@@

\noindent We further classify each word $w \in \T^{*}$ depending on the result of $\parse$:

\begin{definition}
  A word $w \in \T^{*}$ is:
  \begin{itemize}
  \item not in the language if $\parse(w) = \emptyset$.
  \item unambiguous if $\exists t \in T.\ \parse(w) = \{t\}$.
  \item ambiguous if $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$.
  \end{itemize}
\end{definition}

\noindent We can connect these cases to the classical definition of a language:

\begin{itemize}
\item Given a language defined by $\parse$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
\item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
\end{itemize}

\noindent A \emph{resolvably} ambiguous word is a word where all its interpretations can be written in an unambiguous way, or more formally:

\begin{definition}
  A word $w \in \T^{*}$ in a language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-word}
\end{definition}

\noindent The example in Fig.~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way. However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously interpreted as the first and second tree, respectively.

Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, or more formally:

\begin{definition}
  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall w \in \T^{*}.\ \forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-language}
\end{definition}

>>>>

\newcommand{\support}{\mathit{Supp}}
\newcommand{\NT}{V} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{L} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\amb}{\mathit{amb}}

\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}
\newcommand{\regex}{\mathit{Reg}}
\newcommand{\reqpl}{(}
\newcommand{\reqpr}{)}
\newcommand{\reqp}[1]{\reqpl#1\reqpr}
\newcommand{\pospl}{[}
\newcommand{\pospr}{]}
\newcommand{\posp}[1]{\pospl#1\pospr}
\newcommand{\dfa}{\mathit{dfa}} % Go from label to its corresponding dfa
\newcommand{\passthrough}{\mathit{passthrough}} % Go from label to a set of (label x bool), where each element denotes a child production that can be an only child (i.e., no other terminals whatsoever), and the bool states whether it requires parentheses
\newcommand{\nt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs

@@

Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense. We call this meaning an \emph{interpretation} of a word. Note that a single word may have multiple interpretations. As such, we define an \emph{interpreted language} as:

\begin{definition}
  An interpreted language, written $L(\T, T, \parse)$, consists of
\begin{itemize}
\item an alphabet $\T$,
\item a set of interpretations $T$, and
\item a function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
\end{itemize}

  A word $w \in \T^{*}$ is in the language $L(\T, T, \parse)$, written
  $w \in L(\T, T, \parse)$ iff $\parse(w) \neq \emptyset$.
\end{definition}

\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees. In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs shown in Fig.~\ref{fig:arith-example}.

@@

The ambiguity of a word $w \in \T^{*}$ is defined in terms of $\parse$:

\begin{definition}
  A word $w \in L(\T, T, \parse)$ is ambiguous, written $\amb(w)$, iff

  $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$
\end{definition}

\noindent
Note that the above definition implies that a word $w \in L(\T, T,
\parse)$ is not ambiguous (or unambiguous) if $\exists t \in
T.\ \parse(w) = \{t\}$.

We can connect the above definition of an interpreted language to the
classical definition of a language as follows:

\begin{itemize}
\item Given an interpreted language $L(\T, T, \parse)$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
\item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
\end{itemize}

\noindent A \emph{resolvably ambiguous} word is a word where all its interpretations can be written in an unambiguous way, formally:

\begin{definition}\label{def:resolvable-word}
  A word $w \in L(\T, T, \parse)$ is resolvably ambiguous, written $\rho_{\Sigma}(w, \parse)$, iff

  $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$.
\end{definition}

For clarity we abbreviate $\rho_{\T}(w, \parse)$ as $\rho(w)$ if $\T$
and $\parse$ are clear from the context.

The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way. However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously interpreted as the first and second tree, respectively.

Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, formally:

\begin{definition}\label{def:resolvable-language}
  An interpreted language $L(\T, T, \parse)$ is resolvably ambiguous, written $\rho(L)$, iff

  $\forall w \in L(\T, T, \parse).\ \rho_{\T}(w, \parse)$.
\end{definition}
